# ARD-first tables

## ARD-first Tables

Similar to functions that accept a data frame, the package exports functions with nearly identical APIs that accept an ARD.

::: columns
::: {.column width="50%"}

```r
tbl_summary()

tbl_continuous()

tbl_wide_summary()
```

:::
::: {.column width="50%"}

```r
tbl_ard_summary()

tbl_ard_continuous()

tbl_ard_wide_summary()
```

:::
:::

## ARD-first Tables

We can use the skills we learned earlier today to create ARDs for gtsummary tables.

```{r}
#| label: 'ard_stack'
#| output-location: "column"
#| message: true
library(cards)
library(gtsummary)

ard <- ard_stack(
  data = trial, 
  ard_continuous(variables = age),
  ard_categorical(variables = grade),
  .attributes=TRUE
)
ard
```

::: aside
The `.attributes=TRUE` call adds column attributes, like labels, to the ARD table, which leads to better defaults by displaying the column labels in the table.
:::

## ARD-first Tables

We can simply use the ARD from the previous slide, and pass it to `tbl_ard_summary()` for a summary table.

```{r}
#| label: 'tbl_ard_summary'
#| output-location: "column"
#| message: true

tbl_ard_summary(ard)
```


## ARD-first Tables

Now let's try a somewhat more complicated table.

```{r}
#| label: 'tbl_ard_summary2'
#| output-location: "slide"
#| code-line-numbers: "|3|11|12|13"
trial |> 
  ard_stack( 
    .by = trt,
    ard_continuous(variables = age),
    ard_categorical(variables = grade),
    # add this for best-looking tables
    .attributes = TRUE
  ) |> 
  tbl_ard_summary(
    by = trt,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{mean} ({sd})", "{min} - {max}"),
    label = list(age = "Patient Age, yrs")
  ) |> 
  modify_caption("**Table 1. Subject Demographics**")
```

## ARD-first Table *Shells*

```{r}
#| label: 'tbl_ard_summary-mock'
#| output-location: "slide"
#| code-line-numbers: "9-11"
trial |> 
  ard_stack( 
    .by = trt,
    ard_continuous(variables = age),
    ard_categorical(variables = grade),
    # add this for best-looking tables
    .attributes = TRUE
  ) |> 
  update_ard_fmt_fn(stat_names = c("mean", "sd", "min", "max", "p"), 
                    fmt_fn = \(x) "xx.x") |> 
  update_ard_fmt_fn(stat_names = "n", fmt_fn = \(x) "xx") |> 
  tbl_ard_summary(
    by = trt,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{mean} ({sd})", "{min} - {max}"),
    label = list(age = "Patient Age, yrs")
  ) |> 
  modify_header(all_stat_cols() ~ "**{level}**  \nN = xx")
```


## {gtsummary} Exercise 2

1. Navigate to Posit Cloud script `05-gtsummary_exercise2.R`.

2. Create the table outlined in the script.

3. Add the "completed" sticky note to your laptop when complete.

```{r}
#| echo: false
#| cache: false
library(countdown)
countdown(minutes = 10)
```
