# ARD-first tables

## ARD-first Tables

Similar to functions that accept a data frame, the package exports functions with nearly identical APIs that accept an ARD.

::: columns
::: {.column width="50%"}

```r
tbl_summary()

tbl_continuous()

tbl_wide_summary()
```

:::
::: {.column width="50%"}

```r
tbl_ard_summary()

tbl_ard_continuous()

tbl_ard_wide_summary()
```

:::
:::

## ARD-first Tables

We can use the skills we learned earlier today to create ARDs for gtsummary tables.

```{r}
#| label: 'ard_stack'
#| output-location: "column"
#| message: true
library(cards)
library(gtsummary)

ard <- ard_stack(
  data = trial, 
  ard_continuous(variables = age),
  ard_categorical(variables = grade),
  .attributes=TRUE
)

ard |> print(n = 7)
```

::: aside
The `.attributes=TRUE` call adds column attributes, like labels, to the ARD table, which leads to better defaults by displaying the column labels in the table.
:::

## ARD-first Tables

We can simply use the ARD from the previous slide, and pass it to `tbl_ard_summary()` for a summary table.

```{r}
#| label: 'tbl_ard_summary'
#| output-location: "column"
#| message: true

tbl_ard_summary(ard)
```


## ARD-first Tables

Now let's try a somewhat more complicated table.

```{r}
#| label: 'tbl_ard_summary2'
#| output-location: "slide"
#| code-line-numbers: "|3|7-8|10-16|17"
trial |> 
  ard_stack( 
    .by = trt,
    ard_continuous(variables = age),
    ard_categorical(variables = grade),
    # add this for best-looking tables
    .attributes = TRUE,
    .overall = TRUE # get unstratified summary statistics
  ) |> 
  tbl_ard_summary(
    by = trt,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{mean} ({sd})", "{min} - {max}"),
    label = list(age = "Patient Age, yrs"),
    overall = TRUE
  ) |> 
  modify_caption("**Table 1. Subject Demographics**")
```


## What About Other Tables?

::: {.small}

- While our examples have focused on simple demographics tables, the ARD structure is general and any statistic can be presented.

:::

```{r}
#| code-line-numbers: "|2|3-4|5-10|11"
trial |> 
  cardx::ard_stats_t_test_onesample(by = c(trt, grade), variables = age) |> 
  cards::update_ard_fmt_fn(stat_names = "p.value", 
                           fmt_fn = label_style_pvalue(prepend_p = TRUE)) |> 
  tbl_ard_continuous(
    by = trt, 
    include = grade,
    variable = age,
    statistic = ~"{estimate} ({conf.low}, {conf.high}; {p.value})"
  ) |> 
  modify_footnote(all_stat_cols() ~ "One-sample t-test")
```


## {gtsummary} Exercise 2

1. Navigate to Posit Cloud script `exercises/03-gtsummary_partB.R`.

2. Create the table outlined in the script.

```{r}
#| echo: false
#| cache: false
library(countdown)
countdown(minutes = 10)
```

## {gtsummary} Exercise 2 Solution

Create a demographics tables split by TRT01A including AGE, SEX, RACE using ARD-first

```{r}
#| output-location: "slide"
#| code-line-numbers: "|7-14|9|10|11|12|13"
#| message: true
library(cards) 
library(gtsummary) 
adsl <- pharmaverseadam::adsl |> dplyr::filter(SAFFL == "Y") 

# Create the ARD with the needed statistics 
ard <-  
  ard_stack( 
    adsl, 
    .by = TRT01A, 
    ard_continuous(variables = AGE), 
    ard_categorical(variables = c(SEX, RACE)), 
    .attributes = TRUE,  
    .overall = TRUE 
  )
ard
```

## {gtsummary} Exercise 2 Solution

```{r}
#| output-location: "slide"
#| code-line-numbers: "|2-8|3|4|5|6|7"
ard |>  
  tbl_ard_summary( 
    by = TRT01A, 
    include = c(AGE, SEX, RACE), 
    type = AGE ~ 'continuous2', 
    statistic = AGE ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"), 
    overall = TRUE 
  ) 
```


## ARD-first Table *Shells*

```{r}
#| label: 'tbl_ard_summary-mock'
#| output-location: "slide"
#| code-line-numbers: "9-11"
trial |> 
  ard_stack( 
    .by = trt,
    ard_continuous(variables = age),
    ard_categorical(variables = grade),
    # add this for best-looking tables
    .attributes = TRUE
  ) |> 
  update_ard_fmt_fn(stat_names = c("mean", "sd", "min", "max", "p"), 
                    fmt_fn = \(x) "xx.x") |> 
  update_ard_fmt_fn(stat_names = "n", fmt_fn = \(x) "xx") |> 
  tbl_ard_summary(
    by = trt,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c("{mean} ({sd})", "{min} - {max}"),
    label = list(age = "Patient Age, yrs")
  ) |> 
  modify_header(all_stat_cols() ~ "**{level}**  \nN = xx")
```

